Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-05-29 _tex/paper2.tex
Built site for gh-pages

As previously discussed, Case V remains the most widely used model in CJ
literature. This preference largely stems from the widespread adoption
of the BTL model, which provides a simplified statistical representation
of the case. The BTL model mirrors most of Case V's assumptions, with
one notable distinction: while Case V assumes a Normal distribution for
the stimuli' discriminal processes, the BTL model uses the more
mathematically tractable Logistic distribution
\citep{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). However, this substitution has minimal
impact on trait estimation or model interpretation because the scale of
the discriminal process (i.e., the latent trait) is arbitrary up to a
non-monotonic transformation
\citep{vanderLinden_et_al_2017_I, McElreath_2021}. That is, as long as
the substitution (transformation) preserves the data rank order, the
choice of distribution for the discriminal processes is inconsequential.
This condition is satisfied in this case, as the Normal and Logistic
distributions exhibit analogous statistical properties, differing only
by a scaling factor of approximately \(1.7\)
\citep{vanderLinden_et_al_2017_I}.

@@

Notably, while assuming equal dispersions simplifies the trait
measurement model, evidence from the CJ literature suggests that this
assumption may fail to capture the complexity of modern CJ data. In
particular, the assumption may not hold when researchers assess complex
traits or heterogeneous stimuli
\citep{Thurstone_1927a, Bramley_2008, Kelly_et_al_2022}, as these traits
and stimuli can introduce judgment discrepancies due to their unique
characteristics
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
Indeed, the CJ literature may already provide indications of such
discrepancies, particularly in the form of ``misfit'' statistics. For
instance, \emph{misfit texts} are those whose comparisons result in more
judgment discrepancies than those involving other texts
\citep{Pollitt_2004, Pollitt_2012b, Pollitt_2012a, Goossens_et_al_2018}.
These misfit texts may exhibit larger discriminal dispersions or
represent outliers--texts with distinctive characteristics that deviate
markedly from the rest of the sample \citep{Grubbs_1969}. In either
case, the BTL model neither accounts for these cases nor offers any
means to address them, except their elimination from the analysis.

Significant statistical and measurement issues can arise when the
assumption of equal dispersions between stimuli does not hold.
Specifically, the BTL model may overestimate the trait's reliability,
that is, the degree to which the outcome accurately reflects the
\emph{true} discriminal differences between stimuli. This
overestimation, in turn, results in spurious conclusions about these
differences \citep{McElreath_2020} and, by extension, about the
underlying discriminal processes of stimuli. Figure~\ref{fig-dispersion}
also illustrates this scenario when the model's discriminal difference
distribution aligns with the thick continuous line for
\(\sigma_{B}-\sigma_{A}=0\), while the \emph{true} discriminal
difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\). Furthermore, if researchers
acknowledge that misfit statistics may represent texts with different
dispersions or outlying observations, the common CJ practice of
excluding stimuli based on these statistics, as seen in
\citet{Pollitt_2012a}, \citet{Pollitt_2012b},
\citet{vanDaal_et_al_2016}, and \citet{Goossens_et_al_2018}, may
unintentionally discard valuable information, introducing bias into the
trait estimates \citep{Zimmerman_1994, McElreath_2020}. The direction
and magnitude of these biases remain unpredictable, as they depend on
which stimuli researchers exclude from the analysis.

@@

In the first scenario, research on text quality suggests that when
judges evaluate multidimensional, complex traits or heterogeneous
stimuli, they often rely on various intricate characteristics of the
stimuli to form their judgments
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
additional relevant characteristics, when assessed, are unlikely to be
equally weighted or opposing. As a result, they may exert an uneven
influence on judges' perceptions, creating biases that resist
cancellation. For example, this could occur when a judge assessing the
argumentative quality of a text places more weight on its grammatical
accuracy than other judges, thereby favoring texts with fewer errors but
weaker arguments. Furthermore, since the discriminal difference of the
stimuli becomes an observable outcome only through the judges'
perceptions, these biases could introduce dependencies between the
stimuli \citep{vanderLinden_et_al_2017_II}. While direct evidence for
this particular scenario is lacking, studies such as
\citet{Pollitt_et_al_2003} and \citet{vanDaal_et_al_2016} demonstrate
the presence of such biases, supporting the notion that the factors
influencing pairwise comparisons may not always cancel out.

@@

Thus, erroneously assuming zero correlation between stimuli can also
lead to significant statistical and measurement issues. Specifically,
neglecting judges' biases or relevant hierarchical structures can create
dimensional mismatches in the model, leading to the over- or
underestimation of trait reliability
\citep{Ackerman_1989, Hoyle_et_al_2023}. These inaccuracies can result
in spurious conclusions about the discriminal differences
\citep{McElreath_2020} and, by extension, the underlying discriminal
processes of the stimuli. This issue is illustrated in
Figure~\ref{fig-correlation} when the discriminal difference
distribution of the BTL scores follows the thick continuous line
\((\rho = 0)\), while the \emph{true} discriminal difference follows any
discontinuous line where \(\rho \neq 0\).

Finally, as discussed in the previous section, removing \emph{misfit}
judges--that is, judges whose assessments deviate markedly from the
shared consensus
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
and considered outliers under the BTL model \citep{Grubbs_1969}--risk
discarding valuable information and introducing bias into trait
estimates. The direction and magnitude of these biases remain
unpredictable, as they depend on which judges researchers exclude from
the analysis \citep{Zimmerman_1994, OHagan_2018, McElreath_2020}.

@@

Building on the previous section, it is clear that, researchers in CJ
studies typically use the BTL model to measure traits and position the
compared stimuli along a latent continuum \citep{Thurstone_1927b}. The
CJ literature also show that they frequently rely on point estimates of
these traits (i.e, the BTL scores)--or their transformations--to conduct
statistical inference or hypothesis testing. For example, researchers
have used these scores to identify `misfit' judges and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}, detect
biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
calculate correlations with other assessment methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
related to the underlying trait of interest
\citep{Casalicchio_et_al_2015, Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

@@

Overall, the conceptual-population model extends Thurstone's general
form by introducing key innovations to address the limitations discussed
in Section~\ref{sec-theory-issue1b} and Section~\ref{sec-theory-issue2}.
These enhancements include accounting for judges' biases and integrating
hierarchical structural components. Nevertheless, despite its promise of
enhancing measurement accuracy and precision, the model still depends on
the unrealistic assumption that researchers have access to a
\emph{conceptual population}. Since researchers rarely meet this
assumption in practice, they must consider a more realistic scenario.

@@

We can visualize these vectors more clearly using the example in
Equation (\ref{eq-mat13}). Suppose researchers exclude the second
student and the third judge, and they also omit the second text from
each individual and the third repeated judgment from each judge. Given
\(n_{I} = 5\), \(n_{A} = 2\), \(n_{J} = 3\), and \(n_{K} = 3\), these
vectors take the following structure:
\begin{equation}\phantomsection\label{eq-mat24}{
S_{I} = \begin{bmatrix}
1 \\
0 \\
1 \\
1 \\
1
\end{bmatrix} ; \;
S_{J} = \begin{bmatrix}
1 \\
1 \\
0
\end{bmatrix} ; \;
S_{IA} = \begin{bmatrix}
1 \\
0 \\
0 \\
0 \\
1 \\
0 \\
1 \\
0 \\
1 \\
0 
\end{bmatrix} ; \;
S_{JK} = \begin{bmatrix}
1 \\
1 \\
0 \\
1 \\
1 \\
0 \\
0 \\
0 \\
0 
\end{bmatrix}
}\end{equation}

Equation (\ref{eq-mat24}) shows that missing observations in the vectors
\(S_{I}\) and \(S_{J}\)--which represent unsampled students and
judges--directly determine which observations are missing in \(S_{IA}\)
and \(S_{JK}\). In other words, researchers can only observe texts and
judgments from students and judges initially included in the sample. The
equation also shows that the sum of observed elements in \(S_{I}\)
equals the number of sampled students \((n^{s}_{I})\) and that a similar
sum in vector \(S_{J}\) equals the sampled judges \((n^{s}_{J})\).
Conversely, the sum of observed elements in \(S_{IA}\) represents the
total sampled texts across all sampled students
\((n^{s}_{I} \cdot n^{s}_{A})\), while a similar sum in vector
\(S_{JK}\) represents the total sampled repeated judgments across all
sampled judges \((n^{s}_{J} \cdot n^{s}_{K})\). Notice that in this
example, because the design systematically excludes every third repeated
judgment, researchers can also express \(S_{JK}\) using
\(n_{K} = n^{s}_{K} = 2\).

@@

With the definition of \(S\), we incorporate the sample mechanism into
the conceptual-population model. Following the convention of
\citet{McElreath_2020} and \citet{Deffner_et_al_2022},
DAG~\ref{fig-cj14_dag} represents the conceptual-population outcome
\(O^{cp}_{R}\) as unobserved, emphasizing that researchers cannot
directly access this outcome due to the sampling mechanism. The DAG also
depicts the \emph{sample design} vector \(S\) as a causal factor
influencing the sample-comparison outcome \(O^{sc}_{R}\). A square
encloses this vector, indicating that it is a conditioned variable.
Here, \emph{conditioning} means that researchers restrict their focus to
the elements of \(O^{cp}_{R}\) that satisfy \(s_{(i,a,h,b,j,k)}=1\)
\citep{Neal_2020, McElreath_2020}. Essentially, \(S\) serves as the
vector selecting \emph{all repeated judgments a subset of judges makes
for the subset of stimuli produced by the sampled individuals}.

@@

However, due to concerns about the practical feasibility of the
comparison task \citep{Boonen_et_al_2020}, CJ experiments rarely
implement an exhaustive pairings of sampled judges, stimuli, and
individuals. Thus, a realistic scenario must account for the fact that
judges typically compare only a subset of stimuli authored by certain
individuals.

@@

The DAG~\ref{fig-cj14_dag} also integrates the \emph{comparison
mechanism} \(C\) into the conceptual-population model. It shows the
sample-comparison outcome as unobserved, highlighting that researchers
cannot directly access to \(O^{sc}_{R}\) due to the comparison
mechanism. Moreover, the DAG depicts \(C\) as a causal factor
influencing \(O_{R}\), enclosing it in a square to indicate that it is a
conditioned variable. Specifically, \(C\) determines \emph{which
repeated judgments the subset of judges makes for the subset of stimuli
produced by the sampled individuals}. In essence, \(C\) reflects the
assumption that judges \emph{do not} perform all possible repeated
judgments for the selected stimuli. Instead, they carry out a
``sufficient number of comparisons'' \(n_{C}\) to allow the accurate
estimation of the proportion \(P(B>A)\) for each stimulus pair
\citep[pp.~267]{Thurstone_1927b}.

@@

With the structure in SCM~\ref{fig-cj15_scm}, we can derive a
statistical model that addresses violations of the equal dispersion
assumption in CJ experiments (see Section~\ref{sec-theory-issue1a}).
This derivation is possible because a fully specified SCM encodes
functional and probabilistic information, which we can replace with
suitable functions and probabilistic assumptions
\citep{Pearl_et_al_2016}. Specifically, SCM~\ref{fig-cj15_scm} allows us
to express the joint distribution of our complex CJ system as a product
of simpler conditional probability distributions (CPDs)\footnote{This
  re-expression is possible because the \emph{chain rule} of probability
  and the \emph{Bayesian Network Factorization (BNF)} property. For
  further details, see \citet{Pearl_et_al_2016} and \citet{Neal_2020}.},
as shown in Equation (\ref{eq-mat51}). For clarity, we treat expressions
such as \(Y := f_{Y}(X)\), \(P(Y \mid X)\), and \(Y \sim f(Y \mid X)\)
as equivalent, where \(P(Y \mid X)\) and \(f(Y \mid X)\) represent the
CPD of \(Y\) given \(X\).
\begin{equation}\phantomsection\label{eq-mat51}{
\begin{aligned}
  P(O_{R}, & S, C, D_{R}, T_{IA}, X_{IA}, e_{IA}, T_{I}, X_{I}, e_{I}, B_{JK}, Z_{JK}, e_{JK}, B_{J}, Z_{J}, e_{J} ) & \\
  &= P(O_{R} \mid D_{R}, S, C) \cdot P(S) \cdot P(C) \cdot P(D_{R} \mid T_{IA}, B_{JK}) \\
  & \quad \cdot P(T_{IA} \mid T_{I}, X_{IA}, e_{IA}) \cdot P(T_{I} \mid X_{I}, e_{I}) \\
  & \quad \cdot P(B_{JK} \mid B_{J}, Z_{JK}, e_{JK}) \cdot P(B_{J} \mid Z_{J}, e_{J}) \\
  & \quad \cdot P(X_{IA}) \cdot P(X_{I}) \cdot P(Z_{JK}) \cdot P(Z_{J}) \\
  & \quad \cdot P(e_{IA}) \cdot P(e_{I}) \cdot P(e_{JK}) \cdot P(e_{J})  
\end{aligned}
}\end{equation}

Each CPD in Equation (\ref{eq-mat51}) rests on specific assumptions,
which we outline in SCM~\ref{fig-cj16_scm}, the probabilistic model
\ref{fig-cj16_prob}, and the statistical model \ref{fig-cj16_stat}. We
begin by assuming that \(O_{R}\) follows a Bernoulli
distribution\footnote{The binomial distribution--including its special
  case, the Bernoulli distribution--represent a maximum entropy
  distribution for binary events \citep[pp.~34]{McElreath_2020}. This
  means that the Bernoulli distribution is the most consistent
  alternative when only two un-ordered outcomes are possible and their
  expected frequencies are assumed to be constant
  \citep[pp.~310]{McElreath_2020}. For a detailed discussion of the
  binomial as a maximum entropy distribution, see
  \citet[sec.~10.1.2]{McElreath_2020}.}, reflecting the binary nature of
CJ outcomes. Furthermore, following the conventions of Generalized
Linear Models (GLMs)
\citep{Nelder_et_al_1983, Nelder_et_al_1996, Agresti_2015}, the
distribution links \(O_{R}\) to the latent discriminal difference vector
\(D_{R}\) using an inverse-logit function:
\(\text{inv\_logit}(x) = 1/(1 + \exp(-x))\).

@@

\[
\begin{aligned}
  O_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} ) \right] \\
  D_{R} & = \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
  T_{IA} & = T_{I} + \beta_{XA} X_{IA} + e_{IA} \\
  T_{I} & = \beta_{XI} X_{I} + e_{I} \\
  B_{JK} & = B_{J} + \beta_{ZK} Z_{JK} + e_{JK} \\
  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\ \\
  \boldsymbol{e} & \sim \text{Multi-Normal}( \boldsymbol{\mu}, \boldsymbol{\Sigma} )
  \\
  \boldsymbol{\Sigma} &= \boldsymbol{D} \boldsymbol{R} \boldsymbol{D} \\ \\
\end{aligned}
\]

@@

Note that while the joint distribution in Equation (\ref{eq-mat51})
includes the probability distributions of the sampling and comparison
mechanisms, \(P(S)\) and \(P(C)\), as well as those of the predictor
variables--\(P(X_{IA})\), \(P(X_{I})\), \(P(Z_{JK})\), and
\(P(Z_{J})\)--we omit these probabilities from the models presented in
Figure~\ref{fig-cj16}. We make this omission because, although these
distributions contribute to the overall joint distribution of the data,
they do not influence the relationship between \(O_{R}\) and \(D_{R}\).
This independence stems from the random selection designs that govern
\(S\), \(C\), \(X_{IA}\), \(X_{I}\), \(Z_{JK}\), and
\(Z_{J}\)\footnote{Randomization ensures that data--and, by extension,
  an estimator--satisfies several key identification properties, such as
  common support, no interference, and consistency. The most critical
  property, however, is the elimination of confounding.
  \emph{Confounding} occurs when an external variable, such as
  \(X_{I}\), simultaneously influences both the outcome (e.g.,
  \(O_{R}\)) and a variable of interest (e.g., \(S\)), resulting in
  spurious associations between the latter two
  \citep{Everitt_et_al_2010}. Randomization ensure the absence of
  confounding by effectively decoupling the association between the
  variable of interest and any other variable, except for the outcome
  itself. For a more detailed discussion on the benefits of
  randomization, see \citet{Pearl_2009}, \citet{Morgan_et_al_2014},
  \citet{Neal_2020}, and \citet{Hernan_et_al_2020}.}.

>>>>

As previously discussed, Case V remains the most widely used model in CJ
literature. This preference largely stems from the widespread adoption
of the BTL model, which provides a simplified statistical representation
of the case. The BTL model mirrors most of Case V's assumptions, with
one notable distinction. While Case V assumes a Normal distribution for
the stimuli' discriminal processes, the BTL model uses the more
mathematically tractable Logistic distribution
\citep{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). However, this substitution has minimal
impact on trait estimation or model interpretation because the scale of
the discriminal process (i.e., the latent trait) is arbitrary up to a
non-monotonic transformation
\citep{vanderLinden_et_al_2017_I, McElreath_2021}. That is, as long as
the substitution (transformation) preserves the data rank order, the
choice of distribution for the discriminal processes is inconsequential.
This condition is satisfied in this case, as the Normal and Logistic
distributions exhibit analogous statistical properties, differing only
by a scaling factor of approximately \(1.7\)
\citep{vanderLinden_et_al_2017_I}.

@@

Notably, while assuming equal dispersions simplifies the trait
measurement model, evidence from the CJ literature suggests that this
assumption may fail to capture the complexity of some traits or account
for heterogeneous stimuli, such as handwritten texts or English
compositions
\citep{Thurstone_1927a, Andrich_1978, Bramley_2008, Kelly_et_al_2022}.
Indeed, the presence of the so-called \emph{misfit} texts may already
signal these limitations, as these are texts that elicit more judgment
discrepancies than others
\citep{Pollitt_2004, Pollitt_2012b, Pollitt_2012a, Goossens_et_al_2018},
possibly due to larger discriminal dispersions or because they are
genuine outliers--meaning, texts with distinctive characteristics that
deviate markedly from the rest of the sample \citep{Grubbs_1969}. In
either case, the BTL model neither accounts for these anomalies nor
provides tools for addressing them, apart from excluding the
``problematic'' texts from analysis.

Significant statistical and measurement issues can arise when the
assumption of equal dispersions between stimuli does not hold.
Specifically, the BTL model may overestimate the trait's reliability,
that is, the degree to which the outcome accurately reflects the
\emph{true} discriminal differences between stimuli. This
overestimation, in turn, results in spurious conclusions about these
differences \citep{McElreath_2020, Wu_et_al_2022} and, by extension,
about the underlying discriminal processes of stimuli.
Figure~\ref{fig-dispersion} also illustrates this scenario when the
model's discriminal difference distribution aligns with the thick
continuous line for \(\sigma_{B}-\sigma_{A}=0\), while the \emph{true}
discriminal difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\). Furthermore, if researchers
acknowledge that misfit statistics may represent texts with different
dispersions or outlying observations, the common CJ practice of
excluding stimuli based on these statistics, as seen in
\citet{Pollitt_2012a}, \citet{Pollitt_2012b},
\citet{vanDaal_et_al_2016}, and \citet{Goossens_et_al_2018}, may
unintentionally discard valuable information, introducing bias into the
trait estimates \citep{Zimmerman_1994, McElreath_2020}. The direction
and magnitude of these biases remain unpredictable, as they depend on
which stimuli researchers exclude from the analysis.

@@

In the first scenario, research on text quality suggests that when
judges evaluate multidimensional, complex traits or heterogeneous
stimuli, they often rely on a variety of intricate stimulus
characteristics to inform their judgments
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
characteristics, regardless of their relevance, are unlikely to be
equally weighted or to oppose one another across comparisons. As a
result, they may exert a disproportionate influence on judges'
perceptions, generating biases that persist rather than cancel out. For
example, this could occur when a judge assessing the argumentative
quality of a text may place disproportionate emphasis on handwriting
clarity, thereby favoring neatly written texts despite their weaker
arguments. Moreover, because the discriminal process of stimuli becomes
an observable outcome only through the judges' perceptions, these biases
could introduce dependencies between the stimuli
\citep{vanderLinden_et_al_2017_II}. Evidence from traditional marking
contexts confirms a similar phenomenon \citep{Morin_et_al_2018}, and
additional studies
\citep{Pollitt_et_al_2003, vanDaal_et_al_2016, Bartholomew_et_al_2020a}
support the presence of these biases in CJ--reinforcing the argument
that the factors influencing pairwise comparisons do not always cancel
each other out.

@@

Thus, erroneously assuming zero correlation between stimuli can also
lead to significant statistical and measurement issues. Specifically,
neglecting judges' biases or relevant hierarchical structures can create
dimensional mismatches in the model, leading to the over- or
underestimation of trait reliability
\citep{Ackerman_1989, Hoyle_et_al_2023} and even biases
\citep{Wu_et_al_2022}. These inaccuracies can result in spurious
conclusions about the discriminal differences \citep{McElreath_2020}
and, by extension, the underlying discriminal processes of the stimuli.
This issue is illustrated in Figure~\ref{fig-correlation} when the
discriminal difference distribution of the BTL scores follows the thick
continuous line \((\rho = 0)\), while the \emph{true} discriminal
difference follows any discontinuous line where \(\rho \neq 0\).

Finally, as discussed in the previous section, removing \emph{misfit}
judges--that is, judges whose assessments deviate markedly from the
shared consensus
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
and considered outliers under the BTL model
\citep{Grubbs_1969, Wu_et_al_2022}--risk discarding valuable information
and introducing bias into trait estimates. The direction and magnitude
of these biases remain unpredictable, as they depend on which judges
researchers exclude from the analysis
\citep{Zimmerman_1994, OHagan_2018, McElreath_2020}.

@@

Building on the previous section, it is clear that, researchers in CJ
studies typically use the BTL model to measure traits and position the
compared stimuli along a latent continuum \citep{Thurstone_1927b}. The
CJ literature also show that they frequently rely on point estimates of
these traits--typically the BTL scores or its transformations--to
conduct statistical inference or hypothesis testing. For example,
researchers have used these scores to identify `misfit' judges and
stimuli \citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
detect biases in judges' ratings
\citep{Pollitt_et_al_2003, Pollitt_2012b}, calculate correlations with
other assessment methods \citep{Goossens_et_al_2018, Bouwer_et_al_2023},
or test hypotheses related to the underlying trait of interest
\citep{Casalicchio_et_al_2015, Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

@@

Overall, the conceptual-population model extends Thurstone's general
form by introducing key innovations to address the limitations discussed
in Section~\ref{sec-theory-issue1b} and Section~\ref{sec-theory-issue2}.
These enhancements include accounting for judges' biases and integrating
hierarchical structural components. Nevertheless, despite its promise of
enhancing measurement accuracy and precision, the model still depends on
the unrealistic assumption that researchers have access to data from the
\emph{conceptual population}. Since researchers rarely meet this
assumption in practice, they must consider a more realistic scenario.

@@

We can illustrate the structure of these vectors more clearly with an
example. Suppose researchers exclude the second student, the second text
from each student, and the third judge from the setup shown in Equation
(\ref{eq-mat13}). Given \(n_{I} = 5\), \(n_{A} = 2\), \(n_{J} = 3\), and
\(n_{K} = 3\), the resulting vectors would have the following structure:
\begin{equation}\phantomsection\label{eq-mat24}{
S_{I} = \begin{bmatrix}
1 \\
0 \\
1 \\
1 \\
1
\end{bmatrix} ; \;
S_{J} = \begin{bmatrix}
1 \\
1 \\
0
\end{bmatrix} ; \;
S_{IA} = \begin{bmatrix}
1 \\
0 \\
0 \\
0 \\
1 \\
0 \\
1 \\
0 \\
1 \\
0 
\end{bmatrix} ; \;
S_{JK} = \begin{bmatrix}
1 \\
1 \\
1 \\
1 \\
1 \\
1 \\
0 \\
0 \\
0 
\end{bmatrix}
}\end{equation}

Notably, Equation (\ref{eq-mat24}) shows that missing observations in
the vectors \(S_{I}\) and \(S_{J}\)--which represent unsampled students
and judges--directly determine which observations are missing in
\(S_{IA}\) and \(S_{JK}\). In other words, researchers can only observe
texts and judgments from students and judges initially included in the
sample. The equation also shows that the sum of observed elements in
\(S_{I}\) equals the number of sampled students \((n^{s}_{I})\) and that
a similar sum in vector \(S_{J}\) equals the sampled judges
\((n^{s}_{J})\). Conversely, the sum of observed elements in \(S_{IA}\)
represents the total sampled texts across all sampled students
\((n^{s}_{I} \cdot n^{s}_{A})\), while a similar sum in vector
\(S_{JK}\) represents the total sampled repeated judgments across all
sampled judges \((n^{s}_{J} \cdot n^{s}_{K})\). Notice that in this
example, because the design systematically excludes every third repeated
judgment, researchers can also express \(S_{JK}\) using
\(n_{K} = n^{s}_{K} = 2\).

@@

With the definition of \(S\), we incorporate the sample mechanism into
the conceptual-population model. Following the convention of
\citet{McElreath_2020} and \citet{Deffner_et_al_2022},
DAG~\ref{fig-cj14_dag} represents the conceptual-population outcome
\(O^{cp}_{R}\) as unobserved, emphasizing that researchers cannot
directly access this outcome due to the sampling mechanism. The DAG also
depicts the \emph{sample design} vector \(S\) as a causal factor
influencing the sample-comparison outcome \(O^{sc}_{R}\). A square
encloses \(S\), indicating that it is a conditioned variable. In this
context, \emph{conditioning} means that researchers restrict their focus
to the elements of \(O^{cp}_{R}\) that satisfy \(s_{(i,a,h,b,j,k)}=1\)
\citep{Neal_2020, McElreath_2020}. In essence, \(S\) is a vector that
selects \emph{all repeated judgments made by a subset of judges for a
subset of stimuli produced by the sampled individuals}.

@@

However, due to concerns about the practical feasibility of the
comparison task {[}\citet{Boonen_et_al_2020}; {]}, CJ experiments rarely
implement an exhaustive pairings of sampled judges, stimuli, and
individuals. Thus, a realistic scenario must account for the fact that
judges typically compare only a subset of stimuli authored by a sample
of individuals.

@@

The DAG~\ref{fig-cj14_dag} also incorporates the \emph{comparison
mechanism} \(C\) into the conceptual-population model. It shows the
sample-comparison outcome \(O^{sc}_{R}\) as unobserved, emphasizing that
researchers cannot directly access this variable because of the
comparison mechanism. The DAG further shows \(C\) as a conditioned
variable (enclosed in a square) that causally influences the observed
outcome \(O_{R}\). This structure implies that \(C\) determines
\emph{which repeated judgments judges make for the stimuli produced by
the individuals}. In essence, \(C\) reflects the assumption that judges
\emph{do not} perform all possible repeated judgments but instead
complete a sufficient number, \(n_{C}\), to enable the accurate
estimation of the proportion \(P(B>A)\) for each stimulus pair
\citep[pp.~267]{Thurstone_1927b}.

@@

Using the structural causal model (SCM) \ref{fig-cj15_scm}, we can
derive a statistical model that addresses violations of the equal
dispersion assumption (see Section~\ref{sec-theory-issue1a}). This
derivation is possible because a fully specified SCM encodes functional
and probabilistic information, which we can replace with suitable
functions and probabilistic assumptions \citep{Pearl_et_al_2016}.
Specifically, SCM~\ref{fig-cj15_scm} allows us to express the joint
distribution of our complex CJ system as a product of simpler
conditional probability distributions (CPDs)\footnote{This re-expression
  is possible because the \emph{chain rule} of probability and the
  \emph{Bayesian Network Factorization (BNF)} property. For further
  details, see \citet{Pearl_et_al_2016} and \citet{Neal_2020}.}, as
shown in Equation (\ref{eq-mat51}). For clarity, we treat expressions
such as \(Y := f_{Y}(X)\), \(P(Y \mid X)\), and \(Y \sim f(Y \mid X)\)
as equivalent, where \(P(Y \mid X)\) and \(f(Y \mid X)\) represent the
CPD of \(Y\) given \(X\).
\begin{equation}\phantomsection\label{eq-mat51}{
\begin{aligned}
  P(O_{R}, & S, C, D_{R}, T_{IA}, X_{IA}, e_{IA}, T_{I}, X_{I}, e_{I}, B_{JK}, Z_{JK}, e_{JK}, B_{J}, Z_{J}, e_{J} ) & \\
  &= P(O_{R} \mid D_{R}, S, C) \cdot P(S) \cdot P(C) \cdot P(D_{R} \mid T_{IA}, B_{JK}) \\
  & \quad \cdot P(T_{IA} \mid T_{I}, X_{IA}, e_{IA}) \cdot P(T_{I} \mid X_{I}, e_{I}) \\
  & \quad \cdot P(B_{JK} \mid B_{J}, Z_{JK}, e_{JK}) \cdot P(B_{J} \mid Z_{J}, e_{J}) \\
  & \quad \cdot P(X_{IA}) \cdot P(X_{I}) \cdot P(Z_{JK}) \cdot P(Z_{J}) \\
  & \quad \cdot P(e_{IA}) \cdot P(e_{I}) \cdot P(e_{JK}) \cdot P(e_{J})  
\end{aligned}
}\end{equation}

Each CPD in Equation (\ref{eq-mat51}) rests on specific assumptions,
which we outline in the statistical model \ref{fig-cj16_stat}. We begin
by assuming that \(O_{R}\) follows a Bernoulli distribution\footnote{The
  binomial distribution--including its special case, the Bernoulli
  distribution--represent a maximum entropy distribution for binary
  events \citep[pp.~34]{McElreath_2020}. This means that the Bernoulli
  distribution is the most consistent alternative when only two
  un-ordered outcomes are possible and their expected frequencies are
  assumed to be constant \citep[pp.~310]{McElreath_2020}. For a detailed
  discussion of the binomial as a maximum entropy distribution, see
  \citet[sec.~10.1.2]{McElreath_2020}.}, reflecting the binary nature of
CJ outcomes. Furthermore, following the conventions of Generalized
Linear Models (GLMs)
\citep{Nelder_et_al_1983, Nelder_et_al_1996, Agresti_2015}, the
distribution links \(O_{R}\) to the latent discriminal difference vector
\(D_{R}\) using an inverse-logit function:
\(\text{inv\_logit}(x) = 1/(1 + \exp(-x))\).

@@

\[
\begin{aligned}
  O_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} ) \right] \\
  D_{R} & = \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
  T_{IA} & = T_{I} + \beta_{XA} X_{IA} + e_{IA} \\
  T_{I} & = \beta_{XI} X_{I} + e_{I} \\
  B_{JK} & = B_{J} + \beta_{ZK} Z_{JK} + e_{JK} \\
  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\ \\
  \boldsymbol{e} & \sim \text{Multi-Normal}( \boldsymbol{\mu}, \boldsymbol{\Sigma} )
  \\
  \boldsymbol{\Sigma} &= \boldsymbol{V} \boldsymbol{Q} \boldsymbol{V} \\ \\
\end{aligned}
\]

@@

Note that while the joint distribution in Equation (\ref{eq-mat51})
includes the probability distributions of the sampling and comparison
mechanisms, \(P(S)\) and \(P(C)\), as well as those of the predictor
variables--\(P(X_{IA})\), \(P(X_{I})\), \(P(Z_{JK})\), and
\(P(Z_{J})\)--we omit all of these probabilities from the statistical
model \ref{fig-cj16_stat}. This omission is justified because, while
these distributions contribute to the overall joint distribution of the
data, the variables \(S\), \(C\), \(X_{IA}\), \(X_{I}\), \(Z_{JK}\), and
\(Z_{J}\) are observed and independent of any other variable in the
model. As observed variables, they do not require distributional
assumptions in the same way the idiosyncratic errors do. Furthermore,
their independence follows from the underlying random selection
procedures that govern the variables\footnote{Randomization ensures that
  data--and, by extension, an estimator--satisfies several key
  identification properties, such as common support, no interference,
  and consistency. The most critical property, however, is the
  elimination of confounding. \emph{Confounding} occurs when an external
  variable, such as \(X_{I}\), simultaneously influences both the
  outcome (e.g., \(O_{R}\)) and a variable of interest (e.g., \(S\)),
  resulting in spurious associations between the latter two
  \citep{Everitt_et_al_2010}. Randomization ensure the absence of
  confounding by effectively decoupling the association between the
  variable of interest and any other variable, except for the outcome
  itself. For a more detailed discussion on the benefits of
  randomization, see \citet{Pearl_2009}, \citet{Morgan_et_al_2014},
  \citet{Neal_2020}, and \citet{Hernan_et_al_2020}.}.

Next we define \(D_{R}\) as the difference between the discriminal
processes \(T_{IA}[i, a]\) and \(T_{IA}[h, b]\), representing the
underlying written-quality trait of the compared texts, plus the
corresponding repeated judge bias \(B_{JK}[j, k]\). Note that if we
assume that \(B_{JK}[j,k]\) reflects the difference in stimulus-specific
biases, i.e., \(B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]\), we
can re-write the discriminal difference as:
\begin{equation}\phantomsection\label{eq-mat52}{
\begin{aligned}
D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
\end{aligned}
}\end{equation}

This formulation reveals that the discriminal difference captures a
\emph{pure interaction effect}, in which neither the texts' discriminal
processes nor the judges' biases alone determine the outcome, but their
interaction does \citep{Attia_et_al_2022}. Put simply, this mathematical
description captures the idea that the stimuli' discriminal processes
become an observable outcome only through the lens of judges'
perceptions (i.e., their biases). For clarity, the square brackets in
\(D_{R}\) indicate the relevant indices for each trait vector; they do
not imply any subsetting of the data.
